---
title: "Data Analysis of the NOAA Storm Data set to determine Weather Event's Economic and Public Health consequences"
author: "Allan R. Brewer Cappellin"
output: 
  html_document:
    keep_md: true
    toc: true
    toc_depth: 2
---

## Synopsis

This reaseach is aimed to determine the consequence of diferente weather events. The analysis focuses on economical and public health consequences. The data was obtained from a link in the course Reproducible Reaserch in the Coursera Data Science Specialization. Originaly this data comes from the NOAA Database.

## Downloading and Processing the data

Data downloaded from the the web from this url [Data Set](https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2).

```{r Download_Dataset, echo=TRUE}
## download.file(url = "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2", destfile = "~/Desktop/NOAA_StormData.csv.bz2")
```

### Loading Requiered Pacakges

Before loading the data to R and process the data set we must load/install the requiered pacakges for the Analysis.

```{r Require_Packages, echo=TRUE, message=FALSE, warning=FALSE}
if (!require(dplyr)) { install.packages("dplyr") }
if (!require(ggplot2)) { install.packages("ggplot2") }
```

### Load Data into R

After loading the packages we load the data set to the Environment to star the analysis.

```{r Load_Data, echo=TRUE}
stormdata <- read.csv(bzfile("~/Desktop/NOAA_StormData.csv.bz2"), nrow = 1000)
stormdata <- read.csv("~/Desktop/NOAA_StormData.csv", nrows = 1000)
stormdata <- tbl_df(stormdata)
```
