---
title: "Statistical Inference Project Part 1"
author: "Allan R Brewer Cappellin"
date: "May 19, 2015"
output: 
  pdf_document
  html_document:
    keep_md: true
    toc: true
    toc_depth: 3
---

# Simulating an exponential distribution and comparing to the CLT

## Synopsis

In this part of the project we will reporte the simulation of an exponential distribution with a lambda equal to 0.2 for the simulation. We will investigate how the distribution of the mean of 40 exponentials aproximates the theoretical mean of the distribution. Since we are using a lambda equal to 0.2 we know the theoretil mean of the distribution is 5 and the the theoretical variance of the distribution is 5.

## Simuilation and Results

In this section we will write the R code and simulate the 1000 exponential of size 40. This simulation will help answear the questions posted for the project.

### Simulations

The first part of the project is to simulate the 1000 exponential. The following R code is for the simulation.

```{r Simulation, echo=TRUE}
## Set a variable with the number of simulations
nsims <- 1000
## Create an empty vector where the sample means will be stored.
simmeans <- NULL
simsd <- NULL
## We do 1000 simulation where the mean of 40 exponentials are recorded
set.seed(2345) ## We first seet the seed
for (i in 1:nsims) {simmeans <- c(simmeans, mean(rexp(40,0.2)))}
## We do 1000 simulation where the standard deviation of 40 exponentials are recorded
set.seed(2345) ## We set the same seed as before so we work with the same valuessd()
for (i in 1:nsims) {simsd <- c(simsd, sd(rexp(40,0.2)))}
```

### Compare sample mean vs theoretical mean

As i mentioned in the synopsis the theoretical mean for the population is equal 1/lambda that is 1/0.2 = 5. With the vector created in the previous section we can determine the sample mean with a a couple of lines of R code.

```{r sample_mean, echo-TRUE}

```