---
title: "Machine Learning Project"
author: "Allan R Brewer Cappellin"
date: "July 24, 2015"
output:
  pdf_document:
    latex_engine: xelatex
    fig_width: 6
    fig_height: 4
subtitle: Predict the exercise is done using machine learning in R.
fontsize: 10pt
geometry: margin = 1in
---

## Executive Summary

In this project we will train a machine learning model to determine if barbell lifts are done correctly. We must choose the varaibles and algorithim to use. The results must be validated against a testing data setmited to the Coursera Project Submission system.

---

## Exploratory Data Analysis and Summary

```{r Load_Packages, echo=FALSE, message=FALSE, warning=FALSE}
library(caret)
library(rpart)
library(AppliedPredictiveModeling)
library(ElemStatLearn)
```

```{r Load_Data, echo=TRUE}
train <- read.csv(file = "pml-training.csv", stringsAsFactors = FALSE)
test <- read.csv(file = "pml-testing.csv", stringsAsFactors = FALSE)
```


```{r variable_analysis, echo=TRUE}

```



## Machine Learning Model Training


```{r model_training, echo=FALSE}

```


### Final Model


```{r model_summary, echo=FALSE}

```


```{r out_sample_error, echo=FALSE, fig.align='center'}

```

## Conclusions

